---
title: "Inside the Price Tag: Predicting Airbnb Rental Prices in NYC"
author: "Ash Bhuiyan"
date: "05/05/2025"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.width = 7,
  fig.height = 5
)
set.seed(123)
```

### 1. Introduction

My Final Project aims to **predict Airbnb rental prices** in New York City using listing features, geography, host behavior, and guest reviews.
The project is driven by the question: What factors most significantly influence how much a property is listed for on Airbnb?
Using a comprehensive dataset of nearly 49,000 listings from 2019, I explore how variables such as neighborhood, room type, number of reviews, availability, and host activity affect pricing dynamics.

The analysis begins with thorough data cleaning and feature engineering to ensure accurate and interpretable results. Visual exploration then uncovers important trends across boroughs and room types.

Through a combination of visual exploration and machine learning models—including Linear Regression, Decision Tree, and Random Forest—the analysis reveals meaningful patterns and evaluates model performance to better understand the dynamics of short-term rental pricing in NYC.

### 2. Data Loading & Preview

```{r load-data}
library(tidyverse)
library(lubridate)
library(caret)
library(randomForest)
library(rpart)
library(skimr)
library(knitr)
library(viridis)

airbnb <- read_csv("AB_NYC_2019.csv")
glimpse(airbnb)
summary(airbnb$price)
```

### 3. Data Cleaning & Feature Engineering

```{r clean-engineer}
airbnb_clean <- airbnb %>%
  mutate(
    last_review = ymd(last_review),
    reviews_per_month = replace_na(reviews_per_month, 0)
  ) %>%
  filter(!is.na(name), !is.na(host_name)) %>%
  filter(price > 0, price < 1000, minimum_nights <= 365) %>%
  mutate(log_price = log(price))

skim(airbnb_clean %>% select(price, log_price, neighbourhood_group, room_type))
```

### 4. Exploratory Data Analysis

#### 4.1 Price Distribution by Borough

```{r boxplot-borough}
ggplot(airbnb_clean, aes(x = neighbourhood_group, y = price, fill = neighbourhood_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  coord_cartesian(ylim = quantile(airbnb_clean$price, c(0.05, 0.95))) +
  labs(
    x = "Borough",
    y = "Price (USD)",
    title = "Price Distribution by Borough"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### 4.2 Price Distribution by Room Type

```{r boxplot-room}
ggplot(airbnb_clean, aes(x = room_type, y = price, fill = room_type)) +
  geom_boxplot(alpha = 0.7) +
  coord_flip() +
  labs(
    x = "Room Type",
    y = "Price (USD)",
    title = "Price Distribution by Room Type"
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### 4.3 Host Listings vs. Price

```{r host-bar}
airbnb_clean %>%
  group_by(calculated_host_listings_count) %>%
  summarize(avg_price = mean(price), n = n()) %>%
  filter(calculated_host_listings_count <= 10) %>%
  ggplot(aes(x = factor(calculated_host_listings_count), y = avg_price, fill = avg_price)) +
  geom_col(color = "black", width = 0.7) +
  labs(
    x = "Number of Listings per Host",
    y = "Average Price (USD)",
    title = "Host Size vs. Average Price",
    fill = "Avg Price"
  ) +
  scale_fill_gradient(low = "#deebf7", high = "#08519c") +
  theme_minimal()
```

### 5. Modeling

#### 5.1 Linear Regression

```{r lm-model}
lm_mod <- train(
  log_price ~ neighbourhood_group + room_type +
    availability_365 + reviews_per_month + calculated_host_listings_count,
  data = airbnb_clean,
  method = "lm",
  trControl = trainControl(method = "cv", number = 3)
)
lm_mod$results
```

#### 5.2 Decision Tree

```{r rpart-model}
rpart_mod <- train(
  log_price ~ neighbourhood_group + room_type +
    availability_365 + reviews_per_month + calculated_host_listings_count,
  data = airbnb_clean,
  method = "rpart",
  trControl = trainControl(method = "cv", number = 3)
)
rpart_mod$results
```

#### 5.3 Random Forest

```{r rf-model}
set.seed(123)
airbnb_small <- airbnb_clean %>% sample_n(1000)

rf_mod <- randomForest(
  log_price ~ neighbourhood_group + room_type +
    availability_365 + reviews_per_month + calculated_host_listings_count,
  data = airbnb_small,
  ntree = 50,
  mtry = 2
)

rf_mod
```

### 6. Model Comparison Table

```{r compare-table}
model_summary <- data.frame(
  Model = c("Linear Regression", "Decision Tree", "Random Forest"),
  RMSE = c(
    lm_mod$results$RMSE[1],
    rpart_mod$results$RMSE[1],
    sqrt(mean((rf_mod$predicted - airbnb_small$log_price)^2))
  ),
  Rsquared = c(
    lm_mod$results$Rsquared[1],
    rpart_mod$results$Rsquared[1],
    cor(rf_mod$predicted, airbnb_small$log_price)^2
  )
)
kable(model_summary, caption = "Model Performance Summary")
```

### 7. Residual Diagnostics

```{r residuals}
resid <- airbnb_small$log_price - rf_mod$predicted
error_magnitude <- abs(resid)

ggplot(tibble(preds = rf_mod$predicted, resid, error_magnitude), aes(x = preds, y = resid, color = error_magnitude)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  scale_color_viridis(option = "C", direction = -1, name = "Error Magnitude") +
  labs(
    x = "Predicted Log(Price)",
    y = "Residual",
    title = "Residuals vs. Predicted Values"
  ) +
  theme_minimal()
```

### 8. Conclusion

- **Main findings:** Borough, room type, and reviews significantly affect pricing.
- **Best model:** Random Forest performed best using a small sample.
- **Next steps:** Explore full dataset, NLP for reviews, spatial trends.

---